---
title: "fetch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fetch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


In 2020 I wrote a package called [libr](https://libr.r-sassy.org) to deal 
with the problem of managing multiple related datasets.  This package allows
you to manage data from several different types of files:
SAS, Excel, CSV, and R. While the package received generally positive feedback, 
users requested some features that the package did not support:  

* Users wanted to examine the data *before* loading it into memory.
* Users wanted to pick only *some* datasets to load into memory.
* Users wanted to filter all datasets in the library based on a condition.

After consideration of these requests, I determined that these features would
be most easily satisfied by creating a new package.  That package is **fetch**.

## Key Features

The **fetch** package allows you to retrieve data from any of several different 
data sources while minimizing memory usage. 
There are only two steps to retrieve data:

* Create a data catalog
* Fetch data from the catalog

The above two steps are accomplished by the following functions:

* `catalog`: Defines a data catalog for a particular data source
* `fetch`: Pulls a data item from the catalog and loads it into memory

The package has the following key features:

* The **fetch** package loads only the data you need for your analysis.
* The `catalog` function allows you to examine the data before loading it.
* The "filter" parameter on the `catalog` function allows you to filter across 
all datasets in the catalog.
* The `fetch` function loads data, allowing additional subsets if 
you wish.

## How to use **fetch**

Let's see a simple example.  In this example we will create a data catalog,
examine it's contents, and fetch a dataset from the catalog.  First let's create
a data catalog:

```{r eval=FALSE, echo=TRUE}
# Get sample data directory
pkg <- system.file("extdata", package = "fetch")
 
# Create catalog
ct <- catalog(pkg, engines$csv)

# View catalog
ct
# data catalog: 6 items
# - Source: C:/packages/fetch/inst/extdata
# - Engine: csv
# - Items:
    # data item 'ADAE': 56 cols 150 rows
    # data item 'ADEX': 17 cols 348 rows
    # data item 'ADPR': 37 cols 552 rows
    # data item 'ADPSGA': 42 cols 695 rows
    # data item 'ADSL': 56 cols 87 rows
    # data item 'ADVS': 37 cols 3617 rows
```
As can be seen above, this catalog has 6 csv files in it.  The number of columns
and rows are displayed for each file. Information about each file is contained
in the catalog.  This information can be accessed using list notion, like so:

```{r eval=FALSE, echo=TRUE}
# View info for catalog item 
ct$ADEX
# data item 'ADEX': 17 cols 348 rows
# - Engine: csv
# - Size: 70.7 Kb
# - Last Modified: 2020-09-18 14:30:22
#    Name   Column     Class Label Format NAs MaxChar
# 1  ADEX  STUDYID character  <NA>     NA   0       3
# 2  ADEX  USUBJID character  <NA>     NA   0      10
# 3  ADEX   SUBJID character  <NA>     NA   0       3
# 4  ADEX   SITEID character  <NA>     NA   0       2
# 5  ADEX     TRTP character  <NA>     NA   8       5
# 6  ADEX    TRTPN   numeric  <NA>     NA   8       1
# 7  ADEX     TRTA character  <NA>     NA   8       5
# 8  ADEX    TRTAN   numeric  <NA>     NA   8       1
# 9  ADEX   RANDFL character  <NA>     NA   0       1
# 10 ADEX    SAFFL character  <NA>     NA   0       1
# 11 ADEX   MITTFL character  <NA>     NA   0       1
# 12 ADEX  PPROTFL character  <NA>     NA   0       1
# 13 ADEX    PARAM character  <NA>     NA   0      45
# 14 ADEX  PARAMCD character  <NA>     NA   0       8
# 15 ADEX   PARAMN   numeric  <NA>     NA   0       1
# 16 ADEX     AVAL   numeric  <NA>     NA  16       4
# 17 ADEX AVALCAT1 character  <NA>     NA  87      10

```
When the catalog item is printed, it shows a data dictionary for the 
specified dataset.  Column names and data types are displayed, along with
some other useful attributes of your data.

Once the catalog is created, you are ready to fetch data from the catalog:

```{r eval=FALSE, echo=TRUE}

# Fetch data from a catalog
dt <- fetch(ct$ADEX)

# View data
dt
# A tibble: 348 × 17
#    STUDYID USUBJID    SUBJID SITEID TRTP  TRTPN TRTA  TRTAN RANDFL SAFFL MITTFL PPROTFL
#    <chr>   <chr>      <chr>  <chr>  <chr> <dbl> <chr> <dbl> <chr>  <chr> <chr>  <chr>  
#  1 ABC     ABC-01-049 049    01     ARM D     4 ARM D     4 Y      Y     Y      Y      
#  2 ABC     ABC-01-049 049    01     ARM D     4 ARM D     4 Y      Y     Y      Y      
#  3 ABC     ABC-01-049 049    01     ARM D     4 ARM D     4 Y      Y     Y      Y      
#  4 ABC     ABC-01-049 049    01     ARM D     4 ARM D     4 Y      Y     Y      Y      
#  5 ABC     ABC-01-050 050    01     ARM B     2 ARM B     2 Y      Y     Y      Y      
#  6 ABC     ABC-01-050 050    01     ARM B     2 ARM B     2 Y      Y     Y      Y      
#  7 ABC     ABC-01-050 050    01     ARM B     2 ARM B     2 Y      Y     Y      Y      
#  8 ABC     ABC-01-050 050    01     ARM B     2 ARM B     2 Y      Y     Y      Y      
#  9 ABC     ABC-01-051 051    01     ARM A     1 ARM A     1 Y      Y     Y      Y      
# 10 ABC     ABC-01-051 051    01     ARM A     1 ARM A     1 Y      Y     Y      Y      
#  338 more rows
#  5 more variables: PARAM <chr>, PARAMCD <chr>, PARAMN <dbl>, AVAL <dbl>,
#   AVALCAT1 <chr>
#  Use `print(n = ...)` to see more rows

```

At this point, you can proceed with your analysis, or fetch additional datasets 
from the catalog as needed.


## More Examples

Now let's look at some more useful features of the **fetch** package.

### Example 1: Apply a Filter to Fetch Operation


### Example 2: Get Top N Rows


### Example 3: Apply a Pattern to a Catalog

### Example 4: Apply a Filter to a Catalog

### Example 5: Add Import Specifications to a Fetch



### Example 6: Add Import Specifications to a Catalog




### Example 7: Mixing It Up


## Next Steps





